---
title: "Introduction to phylometacheck"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to phylometacheck}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `phylometacheck` package helps microbiome researchers identify and resolve sample name discrepancies between phyloseq objects and metadata files. Using fuzzy matching algorithms, it can detect similar names and suggest corrections.

This vignette demonstrates how to use the package's main functionality.

## Installation

You can install the development version of `phylometacheck` from GitHub:

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("ScottHandley/phylometacheck")
```

## Basic Usage

Let's start by loading the necessary libraries:

```{r setup, message=FALSE, warning=FALSE}
library(phyloseq)
library(phylometacheck)
library(dplyr)
library(ggplot2)
```

For demonstration purposes, we'll use the `GlobalPatterns` dataset from the `phyloseq` package and introduce some sample name inconsistencies:

```{r}
# Load sample data
data(GlobalPatterns)
meta <- data.frame(sample_data(GlobalPatterns))

# Show original sample names
head(rownames(meta), 10)

# Create some deliberate mismatches
set.seed(123)
indices <- sample(1:nrow(meta), 5)
rownames(meta)[indices] <- gsub("\\.", "_", rownames(meta)[indices])
rownames(meta)[indices[1:2]] <- toupper(rownames(meta)[indices[1:2]])
rownames(meta)[indices[3:4]] <- paste0(rownames(meta)[indices[3:4]], "_sample")

# Show modified sample names
head(rownames(meta), 10)
```

### Finding mismatches

Now, let's identify which sample names don't match exactly between the phyloseq object and our modified metadata:

```{r}
# Find mismatches
mismatches <- find_sample_mismatches(GlobalPatterns, meta)

# Show samples only in phyloseq
cat("Samples only in phyloseq:\n")
print(head(mismatches$physeq_only, 10))

# Show samples only in metadata
cat("\nSamples only in metadata:\n")
print(head(mismatches$meta_only, 10))

# Show exact matches
cat("\nNumber of exact matches:", length(mismatches$exact_matches), "\n")
```

### Suggesting fixes with fuzzy matching

Next, we'll use fuzzy matching to suggest possible fixes for the mismatched sample names:

```{r}
# Get suggestions
suggestions <- suggest_name_fixes(GlobalPatterns, meta)

# Show top suggestions
head(suggestions)
```

The suggestions include:
- `physeq_name`: The sample name in the phyloseq object
- `meta_name`: The potential matching sample name in the metadata
- `match_quality`: A score between 0 and 1 indicating the quality of the match
- `patterns`: The patterns of difference that were detected

### Applying fixes

Once we've reviewed the suggestions, we can apply the ones we think are correct:

```{r}
# Select suggestions with high match quality
fixes_to_apply <- suggestions[suggestions$match_quality > 0.8, ]

# Apply fixes to the metadata
fixed_meta <- apply_name_fixes(
  fixes = fixes_to_apply,
  metadata = meta,
  target = "metadata"
)

# Check if the fix was successful
post_fix_mismatches <- find_sample_mismatches(GlobalPatterns, fixed_meta)
cat("Before fixing:", length(mismatches$exact_matches), "exact matches\n")
cat("After fixing:", length(post_fix_mismatches$exact_matches), "exact matches\n")
```

### Visualizing matches and suggestions

We can visualize the relationships between sample names with the built-in plotting function:

```{r, fig.width=8, fig.height=6}
# Plot sample matches and suggestions
plot_sample_matches(mismatches, suggestions)
```

## Interactive HTML Reports

The package can generate comprehensive interactive HTML reports that make it easier to review and fix sample name discrepancies:

```{r eval=FALSE}
# Generate an interactive report
report_path <- render_report(
  physeq = GlobalPatterns,
  metadata = meta,
  output_file = "sample_name_check.html"
)

# Open the report in a browser
browseURL(report_path)
```

The report includes:
- Summary statistics of matches and mismatches
- Interactive tables for exploring the data
- Visualization of sample name relationships
- Suggestions for fixing mismatches
- Code for applying the corrections

## Using the RMarkdown Template

The package also includes an RMarkdown template:

```{r eval=FALSE}
# Create a new report from template
rmarkdown::draft("my_report.Rmd", 
                 template = "phyloseq-meta-check", 
                 package = "phylometacheck")
```

You can customize this template to fit your specific needs.

## Conclusion

The `phylometacheck` package provides a comprehensive solution for identifying and resolving sample name mismatches between phyloseq objects and metadata files. By using fuzzy matching and interactive reports, it makes what used to be a tedious manual process much more efficient and reliable.